---
title: "Environmental Data Loading"
author: "Madrone Environmental Services"
date: "`r Sys.Date()`"
output: html_document
---

# Presence & Absence Data Loading

### Filename:

"002b_EnvironmentalDataLoading_MAFU_V1.Rmd"

### Organization:

Madrone Environmental Services, Ltd.

### Date Created:

March, 1, 2023

### Update:

Formatted according to standards, including final object assignment

### Usage:

Run with R in RStudio, to load your processed environmental parameters (see 002a for Environmental Data Processing) for subsequent use in a SHM to the provincial draft standard. This script is anticipated to follow establishment of presence and absence points (see 001_SpeciesDataLoading.Rmd).

### Inputs:

"ModelData.gdb" "GBIF raw data" "ASTR_range.shp"

### Outputs:

"rast_stack_processed" a process raster stack of all environmental parameters for use in model development.  

Created with R 4.1.1 or later, RStudio 2022.12.0 Build 353 or later (may also run in earlier versions).

# Environmental Data Loading and Stacking

## Purpose

This script is intended to align all environmental predictor layers determined to be relevant for your species, based on your earlier processing steps (see 002a for environmental data processing). Once these data are aligned and stacked, they can be integrated into a landscape level model. 

At the end of this script, users should have a processed raster stack that can be used to develop a species habitat model corresponding to the Provincial draft SHM Standard metadata
documentation and model quality assessment requirements. The raster stack can be saved as an R object as appropriate to limit reloading and re-processing of large files. This option in not included herein. 

When using this framework, the files are managed under the following
folder structure:

```
working directory
  Scripts 
  Data 
  Outputs 
  Report
```
## House-keeping

This chunk establishes a tool for preparing your libraries and setting your
working directory.

### Standardize your libraries for use for the modelling process.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
list.of.packages <- c("tidyverse", "lubridate","chron","bcdata", "bcmaps","sf", "wetlandmapR", "RStoolbox", "doParallel", "rgrass7", "raster", "reproducible")
# Check you have them and load them
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
```

## Map File Loading

This is where we bring in all of the raster layers for environmental variables:

```{r Map Loading}
#Load the DEM - Use this only if your DEM is separate from other parameters, or if you would like to keep it separate or, in this case, if the filename suffix differs from the remaining parameters:
#rast_dem <- raster("R:/22.0253_SHM_Year2/Worked_Examples/Western_Skink/Data/Predictor_Inputs/processed/dem.tif")

#find the folder with your raster layers and pull them out - in this case we haven't processed everything, so two chunks

##Processed:
rastlist_names <- list.files(path = "R:/22.0253_SHM_Year2/Worked_Examples/Coastal_Tailed_Frog/Data/Predictor_Inputs/processed", pattern='_noNA.tif$', 
all.files=TRUE, full.names=TRUE)

#rast_processed <- stack(rastlist_processed,rast_dem)
rast_processed <- stack(rastlist_names)

```




