---
title: "Presence & Absence Data Loading for Coastal Tailed Frog"
author: "Richard Borthwick"
date: "'r Sys.Date()"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Presence & Absence Data Loading

### Filename:

"001_SpeciesDataLoading_V1.Rmd"

### Organization:

Madrone Environmental Services, Ltd.

### Date Created:

March, 1, 2023

### Update:

Formatted according to standards, including final object assignment

### Usage:

Run with R in RStudio, to calibrate your species input data for a SHM to
the provincial draft standard.

### Inputs:

"ModelData.gdb" "GBIF raw data" "ASTR_range.shp"

### Outputs:

"myspecies_input_data.shp" "myspecies_input_data.csv"

Created with R 4.1.1 or later, RStudio 2022.12.0 Build 353 or later (may
also run in earlier versions).

# Species Input Data - Loading and Cleaning

## Purpose

This script is intended to automate data downloads from the Global
Biodiversity Information Framework (GBIF), load additional supporting
data, generate background or absence points for a species, and
spatialize the data for subsequent landscape-level modeling. This is the
first of a 5-script series to produce a standard high-level species
habitat model (SHM).

At the end of this script, users should have a list of objects (see
below) that correspond to the Provincial draft SHM Standard metadata
documentation and model quality assessment requirements, and two file
outputs. The final saved files should be a .csv and a .shp file. The
.csv file should include the location information for species presence
and absence data, as well as pertinent additional information such as
date collected. This same information should be present in the attribute
table of the produced .shp file.

When using this framework, the files are managed under the following
folder structure:

working directory. Scripts Data Outputs Report

Using this same structure within your working directory should allow
mapped pathways to work effectively. If you standardize your folder
structure, most scripts should run fluidly after completing the
housekeeping steps below.

### House-keeping

This chunk establishes a tool for preparing your libraries, setting your
working directory, and establishing manual entries for the Draft SHM
Standard.

```{r Housekeeping, include=FALSE}
#This bit of code turns off warning messages in your R markdown output.
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

## Standardize your libraries for use for the modelling process.
list.of.packages <- c("tidyverse", "lubridate","chron","bcdata", "bcmaps","sf", "rgdal", "readxl", "Cairo", "ggmap","rgbif","stars","raster")
# Check you have them and load them
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

#Set your working directory here:
setwd("R:/22.0253_SHM_Year2/Worked_Examples/Coastal_Tailed_Frog")

#Required Manual Object Entries:

Current_scientific_name <- "Ascaphus truei"

```

## Area of Interest

To commence the modeling process, first, load the area of interest. This
chunk will allow you to load ESRI style feature classes from a
geodatabase. In this example, the species range, which will function as
our Area of Interest, and several species observations are located in
the .gdb, but in different layers. Each layer is addressed
independently.

```{r Loading feature classes in a geodatabase, echo=FALSE}
#OPTIONAL SCRIPT IF NECESSARY LAYERS ARE EMBEDDED IN A GEODATABASE.

fgdb <- file.path("R:/22.0253_SHM_Year2/Worked_Examples/ModelData.gdb")

# List all feature classes in a file geodatabase
fc_list <- ogrListLayers(fgdb)
print(fc_list)

```

```{r Spatiallizing data, echo=TRUE}
#Extract the geodatabase layers of interest:
ASTR_range <- sf::st_read(fgdb, layer = fc_list[[2]])
ASTR_EO <- sf::st_read(fgdb, layer = fc_list[[7]])
ASTR_EO_Ext <- sf::st_read(fgdb, layer = fc_list[[8]])
ASTR_WSI_SO <- sf::st_read(fgdb, layer = fc_list[[9]])
ASTR_iNat_obsc <- sf::st_read(fgdb, layer = fc_list[[6]])

###Alternatively, if a specific file is located in one shape file, carry out the following:

#mymap <- sf::st_read("R:/22.0253_SHM_Year2/Worked_Examples/Western_Skink/Data/Species_Inputs/Raw/R_PLSK_RANGE.shp")

```

## Loading Data from GBIF

Now that the spatial data has been loaded, we can use it to define an
automated data download. Use this script to load data from the Global
Biodiversity Information Framework (GBIF). Please note that it may take
some time if there are a lot of points. Subsequent to loading, we
convert the points to an object of class sf to spatially reference it:

```{r Western Skink Data Loading, echo=TRUE}
gbif_data <- occ_data(scientificName = Current_scientific_name, hasCoordinate = TRUE, limit = 20000,decimalLongitude = "-130, -120", decimalLatitude = "49, 55")

#Note that the extents could be automated from extent(PLSK_range) if your extent is in degrees - ours is UTM, so had to manually enter 

myspecies_coords <- gbif_data$data[ , c("decimalLongitude", "decimalLatitude", "individualCount", "occurrenceStatus", "coordinateUncertaintyInMeters", "institutionCode", "references")]

myspecies_coords <- st_as_sf(myspecies_coords, coords = c("decimalLongitude","decimalLatitude"))

st_crs(myspecies_coords) <- 4326

myspecies_coords <- st_transform(myspecies_coords,3005)
myspecies_coords <- rename(myspecies_coords, "GEOMETRY" = "geometry")


##Optional - if this is your only point data, you can write your shapefile to your desired folder.

######If you need to link multiple datasets or created background data, continue to the next chunk.

# st_write(myspecies_coords,
#          dsn = "R:/22.0253_SHM_Year2/Worked_Examples/Western_Skink/Data/Species_Inputs/Processed",
#          layer = "gbif.shp",
#          driver = "ESRI Shapefile",
#          append = F)

```

### Cleaning Data

There may be points from GBIF that have abundance tallies of zero. This
is a confusing metric, and may need to be addressed based on your
dataset or for your models. The following cleaning scripts can capture
those and identify absence points. These may be important for some
modelling practices, so this process and other cleaning should be
tailored to your dataset and needs.

```{r}
names(myspecies_coords)
sort(unique(myspecies_coords$individualCount))  # notice if some points correspond to zero abundance
sort(unique(myspecies_coords$occurrenceStatus))  # check for different indications of "absent"
```

## Plots

Testing the data and visualizing it:

```{r}
#Plotting AOI as species range and our species points 

ggplot() +
  geom_sf(data = ASTR_range) +
  geom_sf(data = ASTR_EO_Ext) +
  geom_sf(data = ASTR_EO, color = "blue", size = 1) +
  geom_sf(data = ASTR_WSI_SO, color = "green") +
  geom_sf(data = myspecies_coords) 

```

I do this first to do a quick eye-ball test of overlapping points from
different data sets. Note that the CDC Element Occurences are polygons. 

There are multiple options to move forward here. I take the intersections of points and polygons to reduce duplicated data. I rasterize the data so that presence is based on grid size. Only presence data is used for this model.

## Combine the Data

```{r}
### Part 1 - rasterize the EO, so that each cell has one value:
ASTR_EO_rast <- st_rasterize(ASTR_EO %>% dplyr::select(GEOMETRY,OBSERVED_NUMBER))
write_stars(ASTR_EO_rast, "ASTR_EO.tif")
ASTR_EO_rast <- raster("ASTR_EO.tif")


### Part 2 - Ensure that the point data aligns. All obs need a unique ID:

myspecies_coords$SURVEY_OBSERVATION_ID <- c(1:nrow(myspecies_coords))

#join data according to the unique IDs - first confirm projections then combine:

# PLSK_gbif <- st_transform(myspecies_coords, crs=3005)
# PLSK_WSI_SO <- st_transform(PLSK_WSI_SO, crs=3005)

ASTR_gbif <- myspecies_coords
ASTR_gbif <- dplyr::rename(ASTR_gbif, geometry = GEOMETRY)

ASTR_EO_points <- rasterToPoints(ASTR_EO_rast, fun=NULL, spatial=TRUE)
ASTR_sf <- st_as_sf(ASTR_EO_points)
ASTR_sf$SURVEY_OBSERVATION_ID <- c(c(nrow(myspecies_coords)+1):c(nrow(myspecies_coords) + nrow(ASTR_EO_points)))


ASTR_gbif[setdiff(names(ASTR_sf), names(ASTR_gbif))] <- NA
ASTR_sf[setdiff(names(ASTR_gbif), names(ASTR_sf))] <- NA

ASTR_point_map <- rbind(ASTR_sf,ASTR_gbif)
ASTR_point_map$OBSERVED_NUMBER[is.na(ASTR_point_map$individualCount)] = 1
ASTR_point_map$OBSERVED_NUMBER[is.na(ASTR_point_map$OBSERVED_NUMBER)] = 1

ASTR_point_raster <- st_rasterize(ASTR_point_map)#Stars object - change to sf:
final_ASTR_point <- st_as_sf(ASTR_point_raster)
```

## Generate Background or Pseudoabsence Points:

```{r}
ASTR_range <- st_rasterize(ASTR_range %>% dplyr::select(SHAPE))
write_stars(ASTR_range, "ASTR_range.tif")
PLSK_range_rast <- raster("ASTR_range.tif")

background_points <- dismo::randomPoints(PLSK_range_rast,800)
bg_points <- background_points %>%
  as.data.frame %>%
  sf::st_as_sf(coords = c(1,2))
bg_points$SURVEY_OBSERVATION_ID <- c(1:nrow(bg_points)*1000)

#Then we need them to link up to known points:
bg_points <- st_set_crs(bg_points, "EPSG:3005")

bg_points[setdiff(names(ASTR_point_map), names(bg_points))] <- NA
bg_points$OBSERVED_NUMBER[is.na(bg_points$OBSERVED_NUMBER)] = 0

ASTR_points_withbg <- rbind(ASTR_point_map,bg_points)

```

At this stage, background points are randomly dispersed throughout the species range. This was deemed appropriate in this case as the range is well documented, and has previously undergone species specialist review. So, the habitat in that range is prospective habitat. 

## Generate The Data Citations

```{r Citations}
gbif_cit <- gbif_citation(gbif_data)
gbif_cit

```
